Hypercoercions and a Framework for Equivalence of Cast Calculi

### 2019-10-22 paper.tex
casted value -> value in cast

\newcommand{\sOOinspect}[3]{\langle\langle#1,#2\rangle,#3\rangle}
\newcommand{\sOOreturn}[2]{\langle#1,#2\rangle}
\newcommand{\sOOhalt}[1]{\mathtt{Halt} \; #1}

@@

\newcommand{\typingHC}[3]{#1 : #2 \Longrightarrow #3}
\newcommand{\hcvOOinj}[2]{\mathtt{inj} \; #2}
%\newcommand{\hcvOOfun}[5]{\mathtt{fun} \; #1 \; #2 \; #3 \; #4 \; #5}
\newcommand{\hcvOOfun}[5]{\mathtt{fun} \; #2 \; #1 \; #3 \; #4 \; #5}
\newcommand{\hcvOOtt}[0]{\mathtt{unit}}
%\newcommand{\hcvOOcons}[4]{\mathtt{cons}\;#1\;#2\;#3\;#4}
\newcommand{\hcvOOcons}[4]{\mathtt{cons}\;#1\langle#2\rangle\;#3\langle#4\rangle}
%\newcommand{\hcvOOinl}[2]{\mathtt{inl}\;#1\;#2}
%\newcommand{\hcvOOinr}[2]{\mathtt{inr}\;#1\;#2}
\newcommand{\hcvOOinl}[2]{\mathtt{inl}\;#1\langle#2\rangle}
\newcommand{\hcvOOinr}[2]{\mathtt{inr}\;#1\langle#2\rangle}
\newcommand{\hckOOmt}[0]{\mathtt{stop}}
\newcommand{\hckOOconsI}[3]{\mathtt{cons_1}\;#1\;#2\;#3}
\newcommand{\hckOOappII}[2]{\mathtt{app_2}\;#1\;#2}
\newcommand{\sidecond}[1]{\text{if}\;#1}
% Lazy D cast calculus on space-inefficient CEK
\newcommand{\judgeCreduce}[2]{#1 \longmapsto_{\mathcal{C}} #2}
\newcommand{\judgeCreduceTrans}[2]{#1 \longmapsto_{\mathcal{C}}^{*} #2}
\newcommand{\judgeCeval}[2]{eval_{\mathcal{C}}(#1) = #2}
\newcommand{\redrule}[3]{#1 & \longmapsto_\mathcal{C} & #2 & #3\\}
\newcommand{\hiredrule}[3]{\highlight{#1} & \highlight{\longmapsto_\mathcal{C}} & \highlight{#2} & \highlight{#3} \\}
% blame calculus on space-efficient CEK
\newcommand{\judgeSreduce}[3]{#2 \longmapsto_{\mathcal{S}(#1)} #3}
\newcommand{\judgeSreduceTrans}[3]{#2 \longmapsto_{\mathcal{S}(#1)}^{*} #3}
\newcommand{\judgeSeval}[3]{eval_{\mathcal{S}(#1)}(#2) = #3}
\newcommand{\redruleS}[3]{#1 & \longmapsto_{\mathcal{S}(C)} & #2 & #3\\}
\newcommand{\hiredruleS}[3]{\highlight{#1} & 
\highlight{\longmapsto_{\mathcal{S}(C)}} & \highlight{#2} & \highlight{#3} \\}
% Normal Coercion
\newcommand{\ncProj}[2]{#1?^{#2}}
\newcommand{\ncInj}[1]{#1!}
\newcommand{\ncId}[0]{\iota}
\newcommand{\ncSeq}[2]{#1;#2}
\newcommand{\ncFail}[1]{\bot^{#1}}
\newcommand{\ncFun}[2]{\POOfun{#1}{#2}}
\newcommand{\ncProd}[2]{\POOprod{#1}{#2}}
\newcommand{\ncSum}[2]{\POOsum{#1}{#2}}
% Hypercoercion
\newcommand{\hyperCoercionI}[0]{\mathtt{id\star}}
\newcommand{\hyperCoercionC}[3]{#1 \overset{#2}{\curvearrowright} #3}
% machine state simulations
\newcommand{\eqvS}[4]{#3 \approx_{\mathcal{S}\mathcal{S}} #4}
\newcommand{\eqvSD}[3]{#2 \approx_{\mathcal{SD}} #3}
% to-dos
\newcommand{\todoKC}[1]{\todo[inline]{KC needs to #1}}
\newcommand{\todoKCFixed}[0]{\todo[inline]{Fixed. -KC}}
% abbrev
\newcommand{\castCalculus}[0]{$\lambda_{\rightarrow}^{\langle\cdot\rangle}$}
% names
\newcommand{\ineffCEK}{$\mathcal{C}$}
\newcommand{\ineffCEKD}{$\mathcal{D}$}
\newcommand{\ineffCEKUD}{$\mathcal{UD}$}
\newcommand{\judgeDreduce}[2]{#1 \longmapsto_{\mathcal{D}} #2}
\newcommand{\judgeDreduceTrans}[2]{#1 \longmapsto_{\mathcal{D}}^{*} #2}
\newcommand{\judgeDeval}[2]{eval_{\mathcal{D}}(#1) = #2}
\newcommand{\judgeUDreduce}[2]{#1 \longmapsto_{\mathcal{UD}} #2}
\newcommand{\judgeUDreduceTrans}[2]{#1 \longmapsto_{\mathcal{UD}}^{*} #2}
\newcommand{\judgeUDeval}[2]{eval_{\mathcal{UD}}(#1) = #2}
\newcommand{\effCEK}[1]{$\mathcal{S}(#1)$}
\newcommand{\evalEqv}[2]{\ensuremath{eval_{\text{#1}} = eval_{\text{#2}}}}
\newcommand{\continue}[2]{cont(#2,#1)}
\newcommand{\highlight}[1]{{\color{red} #1}}

@@

The transition relation $\judgeCreduce{s}{s}$ is parameterized over
$applyCast$ to allow for the differences between D and UD.
%
When evaluating a cast expression, the machine moves the cast to the
continuation and evaluates the inner expression.
%% Other
%% transition rules starting from$Eval$states are standard.
To apply a casted function, the machine first casts $v_1$, the
operand, then applies the casted operand to $v_2$, the underlying
function, and finally cast the result of the function application.
%
To take out the first (resp. second) part of a casted pair, the
machine firstly take out the first (resp. second) part of $v$, the
underlying pair, and cast the result.
%
%In case splitting, if the target value is a left injection, the
%machine moves to a state that will apply the first continuation
%function to the value inside the left injection. The case for right
%injection is similar.
%
When the target value is a casted sum, the machine moves the cast from
the target value to continuations functions.
%
To cast a value, the machine invokes $applyCast$ on the value.  If the
cast succeeds, the machine returns the result to the next
continuation.  If the cast fails, the machine halts with the blame
label.
%
%% When the continuation is $stop$, the machine enters a halting
%% state. The When the machine is in a halting state, it stays in the
%% same state.

@@

Let $v$ range over values. A value is either the unit, a function (a
closure), a pair, a left injection, a right injection, or a casted
value. As shown by the value typing rules, if the target type of a
casted value is the dynamic type, the underlying value must be of an
injectable type. The definition of injectable types depends on blame
strategies: for the \lazyD\ strategy, every pre-type is injectable;
for the \lazyUD\ strategy, a pre-types is injectable if and only if
all its sub-parts are the dynamic type. If the target type of a casted
value is a pre-type, the type of the underlying value must have the
same type constructor.

@@

Before turning to the definition of the abstract data types, we first
give the definition of values and cast results used by the \effCEK{C}, as they 
are mentioned in the definition of the ADTs.

\todo[inline]{The notation $\hcvOOinj{P}{v}$ is non-standard
   and should be explained. -JS
\\---\\
What is the standard way to display it? -KC}

@@

Let $v$ range over values. In \ineffCEK{} we have one value
constructor for all casted value.  In \effCEK{C}, however, we do not
have this generic value constructor, instead we push those casts into
the ordinary values. For instance, $(\vOOcons{v_1}{v_2})$ corresponds
to $(\hcvOOcons{v_1'}{id(T_1)}{v_2'}{id(T_2)})$. To construct values
of the dynamic type, we use $(\hcvOOinj{P}{v})$.

@@

Fig.~\ref{fig:bisim-SC-D} defines the bisimulation relation.  It is
the smallest congruence relation that also includes the pairs induced
by the rules in Fig.~\ref{fig:bisim-SC-D}.
%
The relation is mutually defined on continuations, written $\kappa
\approx \kappa$, and values, written $v \approx v : T$, where $T$ is
their type.
%
If $k \approx \kappa$, adding an identity cast on top of $k$ gives a related 
continuation.
%
\todo[inline]{The use of ``wrap'' obscures this explanation.
  Consider getting rid of wrap by inlining it. -JS
\\---\\
I agree that inlining wrap and ext can improve readability of $\kappa \approx 
\kappa$, but this would make latex different from Agda, is it alright?
-KC}
%
If $\kappa_1 \approx \kappa_2$, extending $\kappa_1$ and $\kappa_2$ with the 
same cast gives related results.
%
If two values of type $P$ are related, injecting them to $\TOOdyn$ gives 
related values.
%
% The following is part of being a congruence. -JS
%Unit values are of course related. 
%
To relate to a non-casted function in \ineffCEKD, the function in \effCEK{C} 
must use identities as its casts.
%
If functions are related, we can apply the same cast on both side.
%
Rules for products and sums are similar.

>>>>

\newcommand{\sOOinspect}[3]{\langle#1,#2,#3\rangle}
\newcommand{\sOOreturn}[2]{\langle#1,#2\rangle}
\newcommand{\sOOhalt}[1]{\mathtt{Halt} \; #1}

@@

\newcommand{\typingHC}[3]{#1 : #2 \Longrightarrow #3}
\newcommand{\hcvOOinj}[2]{\mathtt{Dyn}_{#1}(#2)}
%\newcommand{\hcvOOfun}[5]{\mathtt{fun} \; #1 \; #2 \; #3 \; #4 \; #5}
\newcommand{\hcvOOfun}[5]{\mathtt{fun} \; #2 \; #1 \; #3 \; #4 \; #5}
\newcommand{\hcvOOtt}[0]{\mathtt{unit}}
%\newcommand{\hcvOOcons}[4]{\mathtt{cons}\;#1\;#2\;#3\;#4}
\newcommand{\hcvOOcons}[4]{\mathtt{cons}\;#1\langle#2\rangle\;#3\langle#4\rangle}
%\newcommand{\hcvOOinl}[2]{\mathtt{inl}\;#1\;#2}
%\newcommand{\hcvOOinr}[2]{\mathtt{inr}\;#1\;#2}
\newcommand{\hcvOOinl}[2]{\mathtt{inl}\;#1\langle#2\rangle}
\newcommand{\hcvOOinr}[2]{\mathtt{inr}\;#1\langle#2\rangle}
\newcommand{\hckOOmt}[0]{\mathtt{stop}}
\newcommand{\hckOOconsI}[3]{\mathtt{cons_1}\;#1\;#2\;#3}
\newcommand{\hckOOappII}[2]{\mathtt{app_2}\;#1\;#2}
\newcommand{\sidecond}[1]{\text{if}\;#1}
% Lazy D cast calculus on space-inefficient CEK
\newcommand{\judgeCreduce}[2]{#1 \longmapsto_{\mathcal{C}} #2}
\newcommand{\judgeCreduceTrans}[2]{#1 \longmapsto_{\mathcal{C}}^{*} #2}
\newcommand{\judgeCeval}[2]{eval_{\mathcal{C}}(#1) = #2}
\newcommand{\redrule}[3]{#1 & \longmapsto_\mathcal{C} & #2 & #3\\}
\newcommand{\hiredrule}[3]{\highlight{#1} & \highlight{\longmapsto_\mathcal{C}} & \highlight{#2} & \highlight{#3} \\}
% blame calculus on space-efficient CEK
\newcommand{\judgeSreduce}[3]{#2 \longmapsto_{\mathcal{S}(#1)} #3}
\newcommand{\judgeSreduceTrans}[3]{#2 \longmapsto_{\mathcal{S}(#1)}^{*} #3}
\newcommand{\judgeSeval}[3]{eval_{\mathcal{S}(#1)}(#2) = #3}
\newcommand{\redruleS}[3]{#1 & \longmapsto_{\mathcal{S}(C)} & #2 & #3\\}
\newcommand{\hiredruleS}[3]{\highlight{#1} & 
\highlight{\longmapsto_{\mathcal{S}(C)}} & \highlight{#2} & \highlight{#3} \\}
% Normal Coercion
\newcommand{\ncProj}[2]{#1?^{#2}}
\newcommand{\ncInj}[1]{#1!}
\newcommand{\ncId}[0]{\iota}
\newcommand{\ncSeq}[2]{#1;#2}
\newcommand{\ncFail}[1]{\bot^{#1}}
\newcommand{\ncFun}[2]{\POOfun{#1}{#2}}
\newcommand{\ncProd}[2]{\POOprod{#1}{#2}}
\newcommand{\ncSum}[2]{\POOsum{#1}{#2}}
% Hypercoercion
\newcommand{\hyperCoercionI}[0]{\mathtt{id\star}}
\newcommand{\hyperCoercionC}[3]{#1 \overset{#2}{\curvearrowright} #3}
% machine state simulations
\newcommand{\eqvS}[4]{#3 \approx_{\mathcal{S}\mathcal{S}} #4}
\newcommand{\eqvSD}[3]{#2 \approx_{\mathcal{SD}} #3}
% to-dos
\newcommand{\todoKC}[1]{\todo[inline]{KC needs to #1}}
\newcommand{\todoKCFixed}[0]{\todo[inline]{Fixed. -KC}}
% abbrev
\newcommand{\castCalculus}[0]{$\lambda_{\rightarrow}^{\langle\cdot\rangle}$}
% names
\newcommand{\ineffCEK}{$\mathcal{C}$}
\newcommand{\ineffCEKD}{$\mathcal{D}$}
\newcommand{\ineffCEKUD}{$\mathcal{UD}$}
\newcommand{\judgeDreduce}[2]{#1 \longmapsto_{\mathcal{D}} #2}
\newcommand{\judgeDreduceTrans}[2]{#1 \longmapsto_{\mathcal{D}}^{*} #2}
\newcommand{\judgeDeval}[2]{eval_{\mathcal{D}}(#1) = #2}
\newcommand{\judgeUDreduce}[2]{#1 \longmapsto_{\mathcal{UD}} #2}
\newcommand{\judgeUDreduceTrans}[2]{#1 \longmapsto_{\mathcal{UD}}^{*} #2}
\newcommand{\judgeUDeval}[2]{eval_{\mathcal{UD}}(#1) = #2}
\newcommand{\effCEK}[1]{$\mathcal{S}(#1)$}
\newcommand{\evalEqv}[2]{\ensuremath{eval_{\text{#1}} = eval_{\text{#2}}}}
\newcommand{\continue}[2]{cont(#2,#1)}
\newcommand{\highlight}[1]{{\color{red} #1}}

@@

The transition relation $\judgeCreduce{s}{s}$ is parameterized over
$applyCast$ to allow for the differences between D and UD.
%
When evaluating a cast expression, the machine moves the cast to the
continuation and evaluates the inner expression.
%% Other
%% transition rules starting from$Eval$states are standard.
To apply a function in cast, the machine first casts $v_1$, the
operand, then applies the operand in cast to $v_2$, the underlying
function, and finally cast the result of the function application.
%
To take out the first (resp. second) part of a pair in cast, the
machine firstly take out the first (resp. second) part of $v$, the
underlying pair, and cast the result.
%
%In case splitting, if the target value is a left injection, the
%machine moves to a state that will apply the first continuation
%function to the value inside the left injection. The case for right
%injection is similar.
%
When the target value is a sum in cast, the machine moves the cast from
the target value to continuations functions.
%
To cast a value, the machine invokes $applyCast$ on the value.  If the
cast succeeds, the machine returns the result to the next
continuation.  If the cast fails, the machine halts with the blame
label.
%
%% When the continuation is $stop$, the machine enters a halting
%% state. The When the machine is in a halting state, it stays in the
%% same state.

@@

Let $v$ range over values. A value is either the unit, a function (a
closure), a pair, a left injection, a right injection, or a value in cast. As 
shown by the value typing rules, if the target type of a
value in cast is the dynamic type, the underlying value must be of an
injectable type. The definition of injectable types depends on blame
strategies: for the \lazyD\ strategy, every pre-type is injectable;
for the \lazyUD\ strategy, a pre-types is injectable if and only if
all its sub-parts are the dynamic type. If the target type of a value in cast 
is a pre-type, the type of the underlying value must have the
same type constructor.

@@

Before turning to the definition of the abstract data types, we first
give the definition of values and cast results used by the \effCEK{C}, as they 
are mentioned in the definition of the ADTs. The notation \hcvOOinj{P}{v} 
stands for values that are cast to the dynamic type.

\todo[inline]{The notation $\hcvOOinj{P}{v}$ is non-standard
   and should be explained. -JS
\\---\\
I switch to the standard notation because we have an observation $\oOOinj$ -KC}

@@

Let $v$ range over values. In \ineffCEK{} we have one value
constructor for all values in casts.  In \effCEK{C}, however, we do not
have this generic value constructor, instead we push those casts into
the ordinary values. For instance, $(\vOOcons{v_1}{v_2})$ corresponds
to $(\hcvOOcons{v_1'}{id(T_1)}{v_2'}{id(T_2)})$. To construct values
of the dynamic type, we use $\hcvOOinj{P}{v}$.

@@

Fig.~\ref{fig:bisim-SC-D} defines the bisimulation relation.  It is
the smallest congruence relation that also includes the pairs induced
by the rules in Fig.~\ref{fig:bisim-SC-D}.
%
The relation is mutually defined on continuations, written $\kappa
\approx \kappa$, and values, written $v \approx v : T$, where $T$ is
their type.
%
If $k \approx \kappa$, adding an identity cast on top of $k$ gives a related 
continuation.
%
\todo[inline]{The use of ``wrap'' obscures this explanation.
  Consider getting rid of wrap by inlining it. -JS
\\---\\
I agree that inlining wrap and ext can improve readability of $\kappa \approx 
\kappa$, but this would make latex different from Agda, is it alright?
-KC}
%
If $\kappa_1 \approx \kappa_2$, extending $\kappa_1$ and $\kappa_2$ with the 
same cast gives related results.
%
If two values of type $P$ are related, injecting them to $\TOOdyn$ gives 
related values.
%
% The following is part of being a congruence. -JS
%Unit values are of course related. 
%
To relate to a naked function in \ineffCEKD, the function in \effCEK{C} 
must use identities as its casts.
%
If functions are related, we can apply the same cast on both side.
%
Rules for products and sums are similar.
